Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-31 paper.tex
Tweaks

To demonstrate the ease and efficiency with this data stored in Zarr
format can be processed on a GPU, we developed two proof-of-concept
applications. In the first we re-implement the af-dist calculation
used throughout as an example of a non-trivial whole-matrix operation
on GPU. In the second, we demonstrate some emerging technology
in which Zarr data can be decompressed directly in GPU memory.
In both cases we demonstrate on chromosome 22 of
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 96,514 variants and 2504 samples, converted to Zarr format
using a chunk size of 10,000 variants $\times$ 1,000 samples.
Experiments in this section were performed on a
computer with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

We first illustrate the ease with which computations can be
be performed on GPU, 
we reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the central operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel, which
is also very similar to the Numba compiled function used in other
experiments. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82 MiB), decompress each to a local
NumPy array and then copy the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with a
data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460 MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer time to 138ms and processing time to 6ms.

Chunk decompression comprises a significant fraction of processing
time (see Figs~\ref{fig-whole-matrix-decode},\ref{fig-s3-throughput})
and host-to-GPU data transfer is
often a bottleneck. It is therefore useful in some applications
to directly decompress Zarr chunks into GPU memory.
The KvikIO~\cite{kvikio} Python package provides a GPU 
implementation of Zarr, such that both data transfers
(file and network IO) and decompression are performed 
directly on the GPU. 
To explore the performance of the GPU optimised compression
codecs available, we recompressed
the \texttt{call\_AD} (allele depth) array
which had dimensions of $96514\times2504\times7$
and an 16 bit integer data type (note that 
we did not use local alleles encoding here, and this 
array is therefore highly sparse and compressible).
For all five codecs considered
(Bitcomp, Snappy, LZ4, Cascaded, Gdeflate)
% GPU Bitcomp compress time = 1.372 (s)
% GPU Snappy compress time = 0.996 (s)
% GPU LZ4 compress time = 0.937 (s)
% GPU Cascaded compress time = 0.703 (s)
% GPU Gdeflate compress time = 0.955 (s)
it took approximately 1s
to compress and store this 3.15 GiB array. 
We then measured the time to decompress these arrays and
copy back to host memory,
% GPU snappy deflate time =  0.606 (s)
% GPU bitcomp deflate time =  0.626 (s)
% GPU cascade deflate time =  0.574 (s)
% GPU gdeflate deflate time =  0.967 (s)
% GPU lz4 deflate time =  0.567 (s)
% CPU zarr time =  7.560 (s)
which varied from 0.57s (Cascaded, LZ4) to 0.97s (Gdeflate) across
the codecs.
For reference, using the Zstandard codec and the
standard CPU Zarr code (which performs some limited multithreading for 
chunk decompression by default) required about 7.6s.

>>>>

To demonstrate the ease and efficiency with data stored in Zarr
format can be processed on a GPU, we developed two proof-of-concept
applications. 
In both cases we demonstrate on data from chromosome 22 of
1000 Genomes WGS data~\citep{byrska2022high}. This consists
of 96,514 variants and 2504 samples, converted to Zarr format
using a chunk size of 10,000 variants $\times$ 1,000 samples.
Experiments in this section were performed on a
computer with an Intel Xeon Silver 4216 processor,
Nvidia A6000 RTX GPU and NVMe SSD.

To demonstate the ease at which non-trivial, novel calculations 
can be performed on a GPU,
we first reimplemented the af-dist calculation
described above using CuPy~\citep{okuta2017cupy}, a GPU accelerated library
of core operations for scientific computing.
All of the operations required to compute  af-dist are
performed on the GPU, mostly with minimal changes required
(e.g.\ replacing \texttt{np.digitize} with \texttt{cp.digitize}).
We perform the central operation of genotype classification using the
CuPy JIT compiler to generate a custom CUDA kernel. 
This is very similar to the Numba compiled function used in other
experiments, within minimal changes required to take advantage of
on-GPU parallelism. In the simplest version, we iterate over the chunks
in the \texttt{call\_genotype} array (stored as 30 chunks, consuming
a total of 7.82 MiB), decompress each to a local
NumPy array and then copy the data to GPU memory by creating
the corresponding CuPy array. For this version, the total time
in the genotype classification kernel was 25ms with a
data transfer time of 179ms (we discount decompression
time here; see next paragraph). For comparison,
our standard Numba compiled classification function required
a total of 1.46s running on a single CPU core.
Another approach in which we load the entire
\texttt{call\_genotype} array into GPU memory at once (possible
here as the $96514 \times 2504 \times 2$ array requires
460 MiB) and uses
a more sophisticated kernel in which we map GPU resources to
subsets of the array in two dimensions,
reduces transfer time to 138ms and processing time to 6ms.

Chunk decompression comprises a significant fraction of processing
time (see Figs~\ref{fig-whole-matrix-decode},\ref{fig-s3-throughput})
and host-to-GPU data transfer is
often a bottleneck. It is therefore useful in some applications
to directly decompress Zarr chunks into GPU memory.
The KvikIO~\cite{kvikio} Python package provides a GPU 
implementation of Zarr, such that both data transfers
(file and network IO) and decompression are performed 
directly on the GPU. 
To explore the performance of the GPU optimised compression
codecs available, we recompressed
the \texttt{call\_AD} (allele depth) array
which had dimensions of $96514\times2504\times7$
and 16 bit integer data type (note that 
we did not use local alleles encoding here, and this 
array is therefore highly sparse and compressible).
For all five codecs considered
(Bitcomp, Snappy, LZ4, Cascaded, Gdeflate)
% GPU Bitcomp compress time = 1.372 (s)
% GPU Snappy compress time = 0.996 (s)
% GPU LZ4 compress time = 0.937 (s)
% GPU Cascaded compress time = 0.703 (s)
% GPU Gdeflate compress time = 0.955 (s)
it took approximately 1s
to compress and store this 3.15 GiB array. 
We then measured the time to decompress these arrays and
copy back to host memory,
% GPU snappy deflate time =  0.606 (s)
% GPU bitcomp deflate time =  0.626 (s)
% GPU cascade deflate time =  0.574 (s)
% GPU gdeflate deflate time =  0.967 (s)
% GPU lz4 deflate time =  0.567 (s)
% CPU zarr time =  7.560 (s)
which varied from 0.57s (Cascaded, LZ4) to 0.97s (Gdeflate) across
the codecs.
For reference, using the Zstandard codec and the
standard CPU Zarr code (which performs some limited multithreading for 
chunk decompression by default) required about 7.6s.
